\textbf{Appendix}\\ Methods for Implementing Integrated Step-Selection Functions with Incomplete Data

### 2024-07-01 Chapter_4/04_Manuscript/Manuscript_Support.tex
Publish PHD

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage{scalerel}
\usepackage{makecell}
\usepackage{ifthen}
\usepackage{tikz}
\usepackage{lineno}
\usepackage{adjustbox}
\usetikzlibrary{svg.path}

@@

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Methods for Implementing Integrated Step-Selection Functions with
Incomplete Data}

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}

Integrated step-selection analyses (iSSAs) are versatile and powerful frameworks
for studying habitat and movement preferences of tracked animals. iSSAs utilize
integrated step-selection functions (iSSFs) to model movements in discrete time,
and thus, require animal location data that are regularly spaced in time.
However, many real-world datasets are incomplete due to tracking devices failing
to locate an individual at one or more scheduled times, leading to slight
irregularities in the duration between consecutive animal locations. To address
this issue, researchers typically only consider bursts of regular data (i.e.,
sequences of locations that are equally spaced in time), thereby reducing the
number of observations used to model movement and habitat selection. We reassess
this practice and explore four alternative approaches that account for temporal
irregularity resulting from missing data. Using a simulation study, we compare
these alternatives to a baseline approach where temporal irregularity is ignored
and demonstrate the potential improvements in model performance that can be
gained by leveraging these additional data. We also showcase these benefits
using a case study on a spotted hyena (\textit{Crocuta crocuta}).

\end{abstract}

%------------------------------------------------------------------------------
%	Main Text
%------------------------------------------------------------------------------
% \newpage
%
% \onehalfspacing
% \tableofcontents
% \doublespacing

% Change page numbering
\newpage
\pagenumbering{arabic}

% Create linenumbers
\linenumbers

\section{Introduction}

Understanding how animals move across the landscape, what habitats they prefer,
and what resources they select are fundamental questions in movement ecology
\citep{Nathan.2008}. Thanks to recent advances in animal tracking
\citep{Cagnacci.2010, Williams.2019, Beardsworth.2022} and remote sensing
technologies \citep{Toth.2016, Rumiano.2020}, new opportunities and analytical
tools have emerged for studying how animals move and interact with their
environment \citep{Tomkiewicz.2010, Kays.2015, Nathan.2022}. Methods commonly
used to analyze animal movement data, including step-selection analyses
\citep{Fortin.2005, Thurfjell.2014, Fieberg.2021} and hidden Markov models
\citep{Michelot.2016}, require \textbf{animal locations} (terms in bold at first
occurrence are defined in \Cref{tab:Glossary}) that are collected at a constant
sampling frequency, leading to data that are equally spaced in time. Yet, it is
common to encounter missing locations in most telemetry data sets
\citep{Frair.2010, Hofman.2019, Vales.2022}, which introduces unwanted
irregularities in the duration between successive locations. Thus, there is a
need for analytical tools that enable the analysis of such data, while
mitigating potential biases arising from temporal irregularity introduced
through missing animal locations.

Step-selection analyses (SSAs) are widely used to study animals' movement
capacities and habitat-selection patterns \citep{Fortin.2005, Thurfjell.2014}.
Straight-line segments connecting consecutive animal locations, referred to as
\textbf{steps}, form the basic building blocks of the statistical likelihood in
SSAs. Specifically, SSAs model the probability $u$ of finding an individual at
location $s$ at time $t+1$, given the animal's past positions at time $t$ and
$t-1$, $s_t$ and $s_{t-1}$, respectively:

\begin{equation}
\label{EQ1}
u(s_{t+1}) = \frac{\phi(s_{t+1}, s_t, s_{t-1}; \gamma)w(x(s_{t+1}); \beta)}{\int_{s \in G}\phi(s_{t+1}, s_{t}, s_{t-1}; \gamma)w(x(s_{t+1}); \beta)ds}
\end{equation}

\noindent Here, the function $\phi$ represents an animal's \textbf{movement
kernel} which is usually expressed in terms of \textbf{step-length} and
\textbf{turning-angle} distributions, with $\gamma$ representing parameters in
these distributions. The function $w$ is the \textbf{habitat-selection function}
and reflects an animal's preferences $\beta$ for environmental characteristics
$x$ at location $s_{t+1}$. In most applications, $w$ is modeled as a log-linear
function of $x$, taking the form $w = exp(x^\top\beta)$. The integral in the
denominator of \ref{EQ1} ensures that $u$ is a proper probability distribution
(i.e., that it integrates to 1). Following \citet{Michelot.2024}, we call the
product $\phi \times w$ the step-selection function (SSF), as it highlights that
the probability of finding an animal at a certain location depends on both the
animal's movement kernel and its habitat-selection function.

Given a series of \textbf{observed steps}, finding the movement and
habitat-selection parameters that maximize the likelihood in \ref{EQ1} requires
approximating the integral in the denominator for each observed step. A variety
of numerical integration techniques can be used for this purpose
\citep{Michelot.2024}, but a common approach is to combine observed steps with
\textbf{random steps} generated by sampling step lengths and turning angles from
parametric distributions informed by the data \citep{Fortin.2005,
Thurfjell.2014}. Environmental conditions at observed steps are then contrasted
with environmental conditions at random steps in a (mixed effects) conditional
logistic regression framework \citep{Fortin.2005, Muff.2020}. To jointly
estimate parameters in $\phi$ and $w$, movement descriptors (e.g., step length
(sl), its natural logarithm (log(sl)), and the cosine of the turning angle
(cos(ta))) can be included in the conditional logistic regression model, and
their estimated coefficients can be used to update the initial (tentative)
step-length and turning-angle distributions \citep{Duchesne.2015, Avgar.2016,
Fieberg.2021}. The specific descriptors that need to be included depend on the
assumed step-length and turning-angle distributions \citep[for more details, see
Appendix C of][]{Fieberg.2021}. This approach to estimating parameters of the
SSF, termed \textit{integrated} SSA (or iSSA) by \citet{Avgar.2016}, is similar
to using importance sampling to approximate the integral in \ref{EQ1}
\citep{Michelot.2024} and is readily accessible through the \texttt{R}-package
\texttt{amt} \citep{Signer.2019}.

%In the SSF framework introduced by \citet{Fortin.2005}, the movement kernel
%$\phi$ was assumed to be known, which precluded joint estimation of $\phi$ and
%$w$ and has been shown to result in biased habitat-selection estimates
%\citep{Forester.2009}. A more robust alternative is to jointly estimate $\phi$
%and $w$ by fitting tentative distributions for step-lengths and turning-angles,
%and including descriptors of step-characteristics (i.e., step-lengths and
%turning-angles) as covariates in the statistical model \citep{Forester.2009,
%Avgar.2016}. In SSFs, the use of random steps serves to generate integration
%points that are needed to approximate the integral in \ref{EQ1} and, even
%though the true movement kernel $\phi$ is often not known a priori, it can be
%approximated using \textit{tentative} distributions \citep{Michelot.2024}. The
%use of parametric distributions serves as a means of conducting importance
%sampling and allows generating random steps that contribute better towards
%improving the approximation \citep{Michelot.2024}.

SSAs have proven extremely effective in numerous ecological studies
\citep{Thurfjell.2014}, providing insights into seasonal space use
\citep{Vales.2022, Enns.2023}, resource selection during distinct behavioral
phases \citep{Elliot.2014, Abrahms.2017, Cozzi.2018, Broekhuis.2019}, and the
effects of landscape familiarity or memory on animal movements \citep{Kim.2023}.
A model parametrized using iSSA resembles a fully mechanistic movement model
that can be used to simulate space use under novel conditions \citep{Avgar.2016,
Signer.2017, Hofmann.2023, Signer.2023}. This characteristic has made iSSAs a
useful tool for quantifying landscape resistance and identifying movement
corridors \citep{Zeller.2020, Buchholtz.2020, Hofmann.2021, Hofmann.2023}.

% Glossary of terms
\input{99_Glossary.tex}

A key assumption when conducting an iSSA is that the data have been collected at
a constant sampling frequency, thus producing \textbf{trajectories} with
\textbf{regular step durations} ($\Delta t$; \citealp{Fortin.2005,
Thurfjell.2014}). Here, we refer to such data as \textbf{regular animal
locations}, and without loss of generality, we assume the regular step duration
to be one (i.e., $\Delta t = 1$). Regular step durations ensure that
step lengths and turning angles are independent of the step duration, and
therefore, steps can be pooled when estimating movement parameters.
Since animal locations are usually obtained using automated tracking devices,
such as GPS collars programmed to record data at regular intervals, satisfying
this assumption seems straightforward. In reality, however, device limitations
often imply that some of the aspired datapoints fail to be collected, thus
introducing \textbf{missingness} and confronting researchers with
\textbf{irregular animal locations} and \textbf{irregular step durations}
\citep{Frair.2010}. In a comprehensive study, \citet{Hofman.2019} showed that
across 3,000 GPS devices and 160 species, the average success rate of obtaining
a scheduled animal location was 78\% (implying a missingess of 22\%), thus
highlighting that irregular animal locations are a frequent phenomenon in
ecological studies.

It is generally recommended that, in the case of such irregular data,
researchers should only retain \textbf{bursts} of steps with regular
step durations (possibly with some tolerance) and discard the rest
\citep{Thurfjell.2014}. We will refer to this modeling approach as having a
\textbf{forgiveness} level of one, indicating that only steps with
step durations $\Delta t = 1$ are retained for further analysis. In \texttt{R},
the \texttt{amt} package provides the function \texttt{track\_resample}
specifically for identifying bursts of steps with regular step durations
\citep{Signer.2019}. The main drawback of this approach is that it may result in
a substantial amount of data being discarded (\Cref{Overview} and
\Cref{NumberOfSteps}). For instance, consider a hypothetical trajectory in which
location 4 is missing (\Cref{Overview}a). The absence of this location prevents
the computation of a step between locations 3 and 4, as well as between
locations 4 and 5. Furthermore, without these steps, it becomes impossible to
compute a turning angle for the step between locations 5 and 6. Consequently,
the lack of a single location reduces the effective sample size, which is the
number of \textbf{valid steps}, by three. Assuming that animal locations are
missing at random, a missingness level of 25\% causes the number of valid steps
to drop by $58\%$ (\Cref{NumberOfSteps}). A modeler willing to increase their
level of forgiveness to two (i.e., allowing for inclusion of steps with $\Delta
t \leq 2$) would be able to increase the number of valid steps by $57\%$
(\Cref{Overview} and \Cref{NumberOfSteps}), therefore achieving a substantial
gain in effective sample size. The ability to capitalize on irregular data is
likely to be particularly important for applications where data are already
limited, such as, for instance, when modeling dispersing individuals
\citep{Rudnick.2012, Fattebert.2015, Cozzi.2020}. However, increasing the
forgiveness also implies that step durations of the retained steps become
irregular, thus necessitating appropriate tools to account for such
irregularity.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Overview.png}
  \caption{(\textbf{a}) Demonstration of how missingness affects the number of
  valid steps that can be used for step-selection analyses under different
  levels of forgiveness. The upper panel depicts a trajectory with zero missing
  locations. That is, all aspired locations were successfully collected on a
  regular interval (yielding a regular step duration of $\Delta t = 1$). This
  trajectory produces four valid steps that can be included in the iSSF model
  and one invalid step that has to be omitted because it has no turning angle
  associated with it. In the central panel, animal location 4 was not obtained,
  introducing a missingness of 16.7\%. If the modeler has a forgiveness of one,
  only a single step can be included for further analysis, as all other steps
  are invalid (either because no turning angle can be computed or because
  step durations exceed the forgiveness). If, however, the modeler exhibited a
  forgiveness of two, such as in the lower panel, a total of three steps could
  be obtained for further analysis. (\textbf{b}) Conceptual illustration of how
  increasing the forgiveness allows one to retain additional steps that can be
  used for step-selection analysis. The sequence of dots resembles the sequence
  of locations that were scheduled to be collected (e.g. using a GPS device),
  with the lines representing hypothetical steps. Because not all locations were
  successfully obtained (gray dots), there is missingness. Depending on the
  forgiveness level, already a single missing location enforces the introduction
  of a new burst, which leads to the loss of several steps. In addition, some of
  the remaining steps are invalid (dotted) because they are lacking a
  turning angle. By increasing the forgiveness, a modeler is willing to retain
  steps that exceed the regular step duration by a certain threshold, which
  enables them to obtain longer bursts and increase the number of steps that can
  be used for further analysis. In the figure, forgiveness increases from left
  to right.}
  \label{Overview}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_NumberOfSteps.png}
  \caption{Illustration of how missingness in animal location data reduces the
  number of valid steps that can be used in step-selection analyses (left panel)
  and how increasing forgiveness helps to retain additional steps that are
  otherwise omitted (right panel). At a missingness of 0, 998 valid steps can be
  obtained from the total of 1,000 animal locations. At higher missingness,
  step durations become irregular, which means that the number of valid steps
  decreases substantially. However, if the modeler is willing to increase their
  forgiveness, additional steps can be gained. The right panel shows the number
  of valid steps that is gained when increasing the forgiveness from 1 to
  2, 3, 4, and 5, respectively. Ribbons indicate the 95\%-percentile intervals
  derived from 1,000 replicates.}
  \label{NumberOfSteps}
  \end{center}
\end{figure}

Various methods have been employed in the past to address temporal
irregularities in animal location data. These may serve as valuable starting
points for developing approaches that enable the integration of irregular data
with iSSFs.

\begin{enumerate}

  \item \textit{Imputation}: An intuitive solution is provided by
  \citet{McClintock.2017}, who suggests fitting a continuous-time correlated
  random walk movement model \citep{Johnson.2008} to the collected data and to
  use the fitted model to impute missing fixes. By imputing missing locations,
  the analysed trajectories become entirely regular again and can be analysed
  using traditional techniques. This approach, which we coin the
  \textit{imputation} approach, is readily available through the
  \texttt{R}-package \texttt{crawl} \citep{Johnson.2022}, yet has only been
  tested for use with hidden Markov movement models and not with iSSFs
  \citep{McClintock.2017}.

  \item \textit{Na\"ive}: Another approach is outlined by \citet{Munden.2021},
  who introduced time-varying iSSFs. In this framework, a change-point detection
  algorithm is applied to the series of observed animal locations to identify
  distinct decision points where the animal turns \citep{Potts.2018,
  Munden.2021}. Steps are then created to represent straight-line movements
  in-between these decision points, but because decision points are not
  regularly spaced in time, the resulting step durations are irregular. Thus,
  step durations are treated as random variables, and, instead of generating
  random steps by sampling step lengths and turning angles, the authors generate
  random steps by sampling step durations, step speeds, and turning angles. The
  underlying assumption is that step lengths scale linearly with step duration,
  and can therefore be meaningfully represented by combinations of step speeds
  and step durations. Although this approach was developed with
  ultra-high-frequency data in mind, we might \textit{na\"ively} apply it more
  broadly to the case of missing data if we believe the presumed
  linear relationship to hold true (this assumption may be reasonable when
  step durations are short, but it is unlikely to hold for longer step durations
  since an animal's path will deviate from a straight line between successive
  locations). Hence, we propose, with our \textit{na\"ive} approach, to scale the
  generated random steps by the observed step duration.

  \item \textit{Dynamic+Model}: Instead of generating random steps by sampling
  step lengths and turning angles from distributions fitted to a single
  step duration, one may choose to fit separate distributions to steps of
  different durations, thus acknowledging potentially non-linear relationships
  between step duration and step lengths or turning angles. Because random steps
  in this approach are sampled using different \textit{tentative} distributions,
  it is necessary to include interactions between step durations and other
  step descriptors (e.g., sl, log(sl), and cos(ta)) in the conditional logistic
  regression model to allow updating the distributions to the different
  step durations. We therefore refer to this approach as the
  \textit{dynamic+model} approach, highlighting that step-length and
  turning-angle distributions are dynamically adjusted to observed
  step durations and that the step duration is included as a modifier of the
  coefficients of step descriptors in the regression model.

  \item \textit{Multistep}: Finally, we propose a \textit{multistep} approach,
  where random steps of varying step durations are generated by stitching
  together sequences of random steps from the regular step duration. For
  example, one can generate a random step of duration $\Delta t = 2$ by
  combining two random steps of step duration $\Delta t = 1$.

\end{enumerate}

\noindent Our goal with this article is to reassess the practice of discarding
irregular animal locations in iSSFs and to investigate whether retaining
irregular data could, in fact, serve to improve model performance. Our
hypothesis is that even irregular data contains valuable information on habitat
and movement preferences that could be leveraged if appropriate methods are
applied. To test this notion, we conducted a comprehensive simulation study
where we simulated regular animal location data with known movement and habitat
parameters. We then introduced varying levels of missingness and applied iSSFs
to estimate simulation parameters. Specifically, we employed the four
alternative iSSF approaches outlined above and compared them to the traditional
approach of including only bursts of regular data and to an uncorrected approach
that simply ignored irregular step durations when using a forgiveness level $>$
1. To examine the impact of different landscape configurations on derived
estimates, we ran our simulations for different levels of spatial
autocorrelation. The use of simulations instead of real data had the benefit
that underlying parameters of the movement kernel and habitat-selection function
were known, which allowed us to assess the reliability of different methods in
retrieving true simulation parameters under different conditions (sensu
\citealp{Kery.2016}). We then compare the traditional approach (using only
bursts of steps with regular step durations) to the best-performing approach
(with irregular data) using GPS locations collected on a spotted hyena
(\textit{Crocuta crocuta}).

We anticipated that increasing forgiveness without adjusting for the introduced
irregularity would entail a bias-variance trade-off. Specifically, we
anticipated that increasing forgiveness would allow improving estimator
precision, but at the cost of introducing bias due to failing to account for
irregular sampling intervals. We expected this bias to be particularly
pronounced at high levels of missingness. Furthermore, we hypothesized that
accounting for irregularity in the \textit{na\"ive}, \textit{dynamic+model}, and
\textit{multistep} approaches would improve model accuracy, while alleviating
potential bias, thus providing an effective means of incorporating additional
data. Because the \textit{imputation} approach relied on an intermediate
movement model to predict missing animal locations, we had no prior expectations
for how well it would perform.

\section{Methods}

We implemented the simulation study in the programming language \texttt{R}
version 4.3.2 \citep{RCoreTeam.2023} and achieved parallelization of
simulation-runs using the \texttt{R}-package \texttt{pbmcapply}
\citep{Kuang.2022}. We generated figures using the \texttt{ggplot2}
\citep{Wickham.2023}, \texttt{ggpubr} \citep{Kassambara.2023}, and
\texttt{ggh4x} \citep{Brand.2023} \texttt{R}-packages. We manipulated raster
data and computed spatial distances using the \texttt{R}-package \texttt{raster}
\citep{Hijmans.2023}. An overview of the simulation design and the different
iSSF approaches is presented in \Cref{Design} and all codes to reproduce this
study are available through an online repository \citep{Hofmann.2023a}.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Design.png}
  \caption{Illustration of the study design. We varied the autocorrelation range
  when simulating spatial covariates from 5 to 20 and tested for different
  missingness scenarios (ranging from 0\% to 50\% missing locations). To
  investigate how increasing forgiveness (i.e., the willingness to include steps
  with duration above the regular step duration) influenced model results, we
  varied its value from 1 (regular step selection) to 5 (considering steps that
  are five times the regular step duration). Finally, we tested five different
  methods to account for potential biases introduced by including irregular
  steps. This gave 3 x 6 x 5 x 5 = 450 combinations, each of which we replicated
  100 times. We assumed step lengths (sl) to follow a gamma distribution,
  whereas turning angles (ta) followed a von Mises distribution.}
  \label{Design}
  \end{center}
\end{figure}

\subsection{Landscape Simulation}

We simulated a virtual landscape comprising two continuous and one categorical
(binary) spatial layers, each with a resolution of 300 x 300 pixels
(\Cref{Covariates}) and spanning across x- and y-coordinates from 0 to 300. The
first layer, \textsf{Dist} (continuous), quantified the distance to the center
of the virtual landscape ($x = 150$, $y = 150$), and can be understood as a
point of attraction, such as, for instance, the center of an animal's
home-range. The second layer, \textsf{Elev} (continuous), resembled an elevation
layer and was simulated by sampling random pixel-values from a normal
distribution. To achieve spatial autocorrelation, we applied a circular moving
window with radius $r$ within which we tallied pixel-values. We varied $r$ from
5, to 10, to 20, depending on the simulated level of autocorrelation (Figure
S1). The third layer, \textsf{Forest} (categorical), represented areas covered
by woodland and was simulated similarly to the \textsf{Elev} layer, but we
binarized the layer by setting all simulated values above the 50\% quantile to
forest and all other values to non-forest (our reference class). We normalized
values of all simulated layers to a range between zero and one and replicated
the simulation of each layer 100 times per autocorrelation scenario, thus
producing 300 unique landscape configurations.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Covariates.png}
  \caption{Example of a landscape configuration across which we simulated
  movement trajectories. All simulated layers had a resolution of 300 x 300
  pixels. The distance layer indicated the distance to the center of the
  landscape and served to simulate attraction. The elevation and forest layers
  were simulated by sampling pixel-values from a normal distribution and
  applying a moving window to achieve spatial autocorrelation. Simulated
  individuals were initiated within the white dashed rectangle, which ensured
  that they would not be released directly at a map border. Simulated
  individuals were attracted to the landscape's center, preferred elevated
  areas, and avoided areas covered by forest. The black line shows the simulated
  trajectory associated with the visualized landscape configuration (cfr.
  \Cref{MovementSimulation}).}
  \label{Covariates}
  \end{center}
\end{figure}

\subsection{Movement Simulation}
\label{MovementSimulation}
To simulate movement across the virtual landscape, we employed the iSSF
simulation algorithm developed by \citet{Signer.2017} and applied in
\citet{Hofmann.2023}. This procedure consists of a sequence of five steps that
are repeated $n$ times to generate a movement trajectory. In step one, we
generated a random starting location by sampling random x- and y-coordinates on
the simulated landscape. To prevent starting points near map borders, we
restricted sampled locations to x- and y-coordinates between 50 and 250 (white
dotted rectangle in \Cref{Covariates}). In step two, we generated a set of 1,000
random steps originating from the current location, by sampling turning angles
from a von Mises distribution with concentration parameter $\kappa = 0.5$ and
step lengths from a gamma distribution with shape parameter $k = 3$ and scale
parameter $\theta = 1$. In step three, we extracted covariate values at the end
of each random step from the underlying covariate layers. In step four, we
assigned to each step $j$ a probability $\pi_j$ of being selected using the
equation below:

\begin{equation}
\label{EQ2}
  \pi_j =
  \frac{exp(\beta^\top x_j)}{\sum_{i = 1}^J exp(\beta^\top x_i)} %=
  % \frac{
  %   exp(\beta_{dist} dist_j + \beta_{elev} elev_j + \beta_{forest} forest_j)
  % }{
  %   \sum_{i = 1}^J exp(\beta_{dist} dist_i + \beta_{elev} elev_i + \beta_{forest} forest_i)
  % }
\end{equation}

\noindent Here, $\beta$ represents the vector of habitat-selection parameters
and $x_j$ the covariate value at the end of the $j$\textsuperscript{th} step.
The probability of a step being selected thus depended on its associated
covariate values, the covariate values of all other random steps, and the
simulated preferences $\beta$. We defined the habitat-selection parameters as
$\beta_{dist} = -20$, $\beta_{elev} = 0.5$, and $\beta_{forest} = -1$. That is,
simulated individuals were attracted to the landscape's center, preferred
elevated areas, and avoided areas covered by forest. In step five, we sampled
one of the random steps based on predicted probabilities and computed the
simulated individual's new position. We then repeated steps two through five
until the trajectory comprised a total of 1,000 steps. Each simulated step was
assumed to have a step duration of exactly $\Delta t = 1$. We repeated the
simulation for each of the 300 simulated landscapes, producing 300 unique
trajectories (example trajectory presented in \Cref{Covariates}).

\subsection{Data Rarefication}

To simulate missingness, we rarefied the trajectories by randomly removing a
fixed fraction of animal locations. To assess the impact of different degrees of
missingness, we varied the fraction of removed data from 0\% (complete dataset)
to 50\% in increments of 10\%. The random removal of animal locations introduced
temporal irregularity, such that the resulting step durations differed depending
on the time elapsed between remaining fixes. We replicated the rarefication of
each trajectory 100 times.

\subsection{Computing Bursts}

We used the rarefied data to compute bursts consisting of a sequence of
animal locations with step durations that did not exceed the forgiveness value.
To test how different levels of forgiveness impacted our results, we varied
forgiveness from 1 (maximum allowed step duration was $\Delta t = 1$) to 5
(maximum allowed step duration was $\Delta t = 5$). As an example, if the
forgiveness was 1, any step with step duration $\Delta t > 1$ resulted in a new
burst. If the forgiveness was 2, in contrast, step durations of up to $\Delta t
= 2$ were allowed before a new burst was introduced (\Cref{Overview}b). Within
each burst, we calculated step lengths and turning angles. However, due to the
grouping of steps into bursts, the orientation of the first step within each
burst relative to the previous step could not be determined. As a result, this
step always lacked a turning angle and was considered invalid
(\Cref{Overview}b).

\subsection{Fitting Distributions}

Based on the steps retained within bursts, we parametrized tentative step-length
and turning-angle distributions. Specifically, we used the \texttt{fit\_distr}
function from the \texttt{amt} package \citep{Signer.2019}, which is a wrapper
function for the \texttt{fitdist} function from the \texttt{fitdistrplus}
package \citep{Delignette-Muller.2015}, and fitted a gamma distribution to
step lengths and a von Mises distribution to turning angles. Notably, we
employed two different fitting procedures:

\begin{enumerate}

  \item \textit{Regular Distributions:} In this procedure, we fitted parametric
  distributions considering only step lengths and turning angles from steps that
  exhibited a step duration of $\Delta t = 1$ (i.e., the regular step duration).
  Any steps with irregular step durations ($\Delta t > 1$) were discarded and
  did not affect distributional parameter estimates. This represents the
  traditional procedure in iSSFs where only regular bursts of animal locations
  are considered when estimating tentative movement parameters.

  \item \textit{Dynamic Distributions:} In this procedure, we fitted separate
  parametric distributions to step lengths and turning angles from steps of
  different step durations. That is, we parametrized separate turning-angle and
  step-length distributions representative of steps with durations of $\Delta t
  = 1, 2, 3, 4$ and $5$ (which corresponds to the maximum forgiveness level we
  tested for). Some step durations only rarely occurred at low levels of
  missingness, thus complicating parametrization of the associated
  distributions. To facilitate estimation of dynamic distribution parameters
  across all $\Delta t$ (Figure S2), we resampled data to different step
  durations using the \texttt{track\_resample} function from the \texttt{amt}
  package \citep{Signer.2019} before fitting tentative parameters. This ensured
  a sufficient number of steps for each step duration to estimate associated
  parameters. An alternative approach would be to increase missingness in the
  data even further, thus introducing a larger number of longer step durations.

\end{enumerate}

\subsection{Step-Selection Functions}

We implemented a baseline \textit{uncorrected} approach and four alternative
iSSF approaches that mainly differed in the way in which random steps were
generated, but sometimes also in the model call that was used to estimate
parameters (\Cref{Design}). In the \textit{uncorrected} approach, we treated
data as if they were regular, ignoring potential issues arising from having
variable step durations. When forgiveness was one, this approach corresponded
to the traditional iSSF approach. All other approaches were targeted towards
reducing potential biases arising from the inclusion of steps with irregular
step durations. Irrespective of the approach employed, we paired each observed
step with a total of 200 random steps:

\begin{itemize}

  \item \textit{Uncorrected}: In the uncorrected approach, we generated random
  steps by sampling step lengths and turning angles from statistical
  distributions fitted to steps with step durations of $\Delta t = 1$,
  regardless of the forgiveness value or observed step durations. Thus, this
  approach ignored any potential effect of step duration when generating random
  step lengths and turning angles.

  \item \textit{Imputed}: In this approach, we sampled step lengths and turning
  angles from statistical distributions fitted to observed steps with a step
  duration of $\Delta t = 1$. However, prior to generating random steps, we
  imputed missing fixes using predictions from a simple movement model.
  Specifically, we fitted a single-state movement model \citep{Johnson.2008} to
  the simulated trajectories and used the parametrized model to predict
  coordinates for all missing animal locations. For this, we used the functions
  \texttt{crwFit} and \texttt{crwPredict} from the \texttt{crawl}
  \texttt{R}-package \citep{Johnson.2022}. Although the \texttt{crwFit} function
  provides capabilities to incorporate location measurement error, we assumed
  animal locations were measured without error. The imputation resulted in a
  complete dataset without any missing animal locations, such that each
  trajectory consisted of a single continuous burst of locations equally spaced
  in time. As such, the imputation approach is not affected by the forgiveness
  level.

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
% \usepackage[nomarkers, nolists]{endfloat}
\usepackage{scalerel}
\usepackage{makecell}
\usepackage{ifthen}
\usepackage{tikz}
\usepackage{lineno}
\usepackage{adjustbox}
\usetikzlibrary{svg.path}

@@

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{\textbf{Appendix}\\ Methods for Implementing Integrated Step-Selection
Functions with Incomplete Data}

@@

%------------------------------------------------------------------------------
%	Main Text
%------------------------------------------------------------------------------
\newpage

% Change page numbering
\pagenumbering{arabic}

% Create linenumbers
% \linenumbers

% Change to appendix counters
\appendix
\beginappendix

%------------------------------------------------------------------------------
%	Appendix S1:
%------------------------------------------------------------------------------
\newpage
\section{Landscape Simulation: Different Autocorrelation Scenarios}
\begin{figure}[!ht]
  \begin{center}
  \includegraphics[width = \textwidth]{99_CovariatesAppendix.png}
  \caption{Simulated landscapes under different levels of autocorrelation (5,
  10, 20; from top to bottom). Autocorrelation only affected the layers
  \textsf{elev} and \textsf{forest}, which were both simulated using a Gaussian
  random field neutral landscape model \citep{Schlather.2015} using the
  \texttt{R}-package \texttt{NLMR} \citep{Sciaini.2018}. Simulations were
  repeated 100 times for each autocorrelation scenario, thus resulting in 300
  unique landscape configurations.}
  \label{Covariates}
  \end{center}
\end{figure}

%------------------------------------------------------------------------------
%	Appendix S2:
%------------------------------------------------------------------------------
\newpage
\section{Dynamic Tentative Distribution Parameters}
\begin{figure}[!ht]
  \begin{center}
  \includegraphics[width = 0.9\textwidth]{99_DistributionParameters.png}
  \caption{Tentative prameter estimates for the von Mises distribution (top row) and
  gamma distribution (bottom row) fitted to steps with different durations. The
  von Mises distribution requires one parameter, namely a concentration
  parameter ($\kappa$). The gamma distribution requires a shape parameter ($k$)
  and a scale parameter ($\theta$). The subscript $_0$ is used to indicate that
  these are tentative distribution parameters (sensu \citealp{Avgar.2016} and
  \citealp{Fieberg.2021}).}
  \label{DistributionParameters}
  \end{center}
\end{figure}

%------------------------------------------------------------------------------
%	Appendix S3:
%------------------------------------------------------------------------------
\newpage
\section{Distribution of Relative Turning Angles following Different Step Durations}
\begin{figure}[!ht]
  \begin{center}
  \includegraphics[width = 0.9\textwidth]{99_TurningAnglePreviousDuration.png}
  \caption{Density of relative turning angles associated with steps of $\Delta t
  = 1$ following steps with different durations for all three autocorrelation
  scenarios (5, 10, and 20). To generate this figure, we assumed a missingness
  of 0.5 and forgiveness of 5.}
  \label{PreviousDuration}
  \end{center}
\end{figure}

%------------------------------------------------------------------------------
%	Appendix S4:
%------------------------------------------------------------------------------
\newpage
\section{Model Estimates across all Scenarios}
\begin{figure}[!ht]
  \begin{center}
  \includegraphics[width = \textwidth]{99_ModelComparisonAppendix.png}
  \caption{Parameter estimates across different autocorrelation scenarios (5,
  10, 20; panels a, b, and c) and missingness levels (0\% - 50\%; from left
  to right). True simulation parameters are indicated by the solid black
  lines. Parameter estimates from the different approaches are given by the
  colored symbols, and their bootstrap 95\% CIs across 100 replicates
  by the colored lines.}
  \label{ModelComparison}
  \end{center}
\end{figure}

%------------------------------------------------------------------------------
%	Appendix S5:
%------------------------------------------------------------------------------
\newpage
\section{Case Study Covariates}

\begin{figure}[htpb]
  \begin{center}
  \includegraphics[width = \textwidth]{99_CaseStudyCovariates.png}
  \caption{Covariates used for the case study, overlaid with the GPS data of a
  spotted hyena called ``Apollo'' (lines in black). Apollo was originally
  collared in 2007 in northern Botswana and monitored until 2011. The depicted
  area is part of the Okavango Delta, which is a massive wetland area. Data was
  projected to a local projection (EPSG:32734).}
  \label{CaseStudyCovariates}
  \end{center}
\end{figure}

%------------------------------------------------------------------------------
%	Appendix S6:
%------------------------------------------------------------------------------
\newpage
\section{Case Study Model Output}

\begin{table}[htpb]
  \begin{center}
  \caption{Model results from the case study using GPS data collected on Apollo.
  In F1, forgiveness was set to one (only 2-hour steps were considered), whereas
  in F3-S and F3-SH a forgiveness of three was employed (allowing for step
  durations of up to 6 hours). In model F3-S, the step duration was interacted
  with step descriptors. In model F3-SH, step duration was interacted with step
  descriptors and habitat covariates.}
  \label{CaseStudyTable}
    \begin{adjustbox}{max width=0.70\textwidth}
    \begin{threeparttable}[h]
      \input{99_CaseStudyTable.tex}
      \begin{tablenotes}
       \item \textit{Significance codes: * \(p < 0.10\) \quad ** \(p < 0.05\)
        \quad *** \(p < 0.01\)}
      \end{tablenotes}
    \end{threeparttable}
  \end{adjustbox}
  \end{center}
\end{table}
